

\newglossaryentry{node}{
	name={NodeJS},
	description={\enquote{NodeJS is a JavaScript runtime that helps with creating scalable applications. With the help of the package manager npm it is also widely used to create web-based applications.} \cite{node}}
}

\newglossaryentry{electron}{
	name={Electron},
	description={\enquote{The framework Electron is used to build cross platform applications with the help of JavaScript, HTML and CSS.} \cite{electron}}
}

\newglossaryentry{react:native}{
	name={React Native},
	description={\enquote{The framework React Native is used to create native applications for iOS and Android with the React UI library.} \cite{react:native}}
}

\newglossaryentry{java:applet}{
	name={Java-Applet},
	description={\enquote{A Java-Applet is small application written in the Java programming language that can be used on the web.} \cite{wiki:javaapplet}}
}

\newglossaryentry{dom}{
	name={DOM},
	description={\enquote{The Document Object Model (DOM) defines a structure of a HTML and XML document. It also regulates how documents can be accessed and manipulated by JavaScript.} \cite{dom}}
}

\newglossaryentry{llvm}{
	name={LLVM},
	description={\enquote{LLVM can be seen as a library for programmatically creating machine-native code. It helps defining new programming languages and enhances development on existing ones.} \cite{llvm:explanation, wasm:build}}
}

\newglossaryentry{mozilla}{
	name={Mozilla},
	description={\enquote{The Mozilla Foundation is a non-for-profit organisation known mostly for the Firefox browser.}}
}

\newglossaryentry{stackmachine}{
	name={Stack machine},
	text={stack-based},
	description={\enquote{A stack machine is a type of a computer that uses a last-in, first-out stack to hold temporary values. Most of its instructions assume that operands will be from the stack, and results placed in the stack.} \cite{wiki:stackmachine}}
}

\newglossaryentry{pointer}{
	name={Pointer},
	text={pointer},
	description={\enquote{A pointer is an object that stores the reference, a memory address, to another value located in the computer memory.} \cite{wiki:pointer}}
}

\newglossaryentry{nodemodule}{
	name={Node Module},
	description={\enquote{A Node Module is a package that consist of reusable code that can be installed through a package manager for usage.}}
}

\newglossaryentry{rust:ownership}{
	name={Ownership},
	text={ownership},
	description={\enquote{Ownership is one of Rust central features. Each value has an owner. When the owner leaves the scope, the value is dropped out of memory.} \cite{rust:book}}
}

\newglossaryentry{markdown}{
	name={Markdown},
	description={\enquote{Markdown is a simple solution for formatting text, that can be converted to HTML.} \cite{wiki:markdown}}
}




